Speeding up prediction performance of the BDT-based models

### 2017-11-11 acat_paper/paper.tex
Added motivation about LHCb

\begin{abstract}
The outcome of a machine learning algorithm is a prediction model. It is quite usual such models are computationaly expensive and growth of the quality of such models leads to the deterioration in the inference speed. However it is not always tradeoff between quality vs speed. In this paper we show it is possible to speed up model by using additional memory without loosing prediction quality significantly for a novel boosted trees algorithm called Catboost. The idea is to combine two approaches: training fewer trees and merging trees into kind of hash maps (DecisionTensors). The proposed method allows for pareto-optimal reduction of the computational complexity of the decision tree model with regard to the quality of the model. In the considered example number of lookups was decreased from 5000 to only 6 (speedup factor of 1000) while AUC score of the model was reduced by less than per mil.
\end{abstract}

@@

In this paper, we consider a modification of the training process of a predictive models by decision tree-based algorithm CatBoost [1]. But our approach is applicable to other boosting decision tree-based algorithms.

@@

CatBoost is the implementation of a tree-based gradient boosting in which oblivious decision trees are used. The values of each feature $i$ are discretized into baskets using the boundaries $B_i={b_{i1},...,b_{ip_i}}$, where $p_i$ is the number of boundaries. The boundaries themselves are determined in advance using various approaches.
Original feature vector with real values is replaced by a vector of binary values $0$ and $1$ ($false$ and $true$). The value of the $i$-th attribute $f_i$ is replaced by a binary vector $G_i={g_{i1},...,g_{ip_i}}$ where $g_{ij} = f_i > b_{ij}$. Finally, all $G_i$ are concatenated into one vector. Consecutively, the resuling decision will a make bunch of boolean checks instead of comparing some real feature value with the thresholds.

@@

\section*{References}
\numrefs{99}
\item Gulin A, Kuralenok I, Pavlov D 2011 {\it Winning The Transfer Learning Track of Yahoo!'s Learning To Rank Challenge with YetiRank.} (Yahoo! Learning to Rank Challenge) p 63--76
\item Likhomanenko T, Ilten P, Khairullin E, Rogozhnikov A, Ustyuzhanin A, Williams M 2015 {\it LHCb Topological Trigger Reoptimization} (Journal of Physics: Conference Series)
\endnumrefs

>>>>

\begin{abstract}
The outcome of a machine learning algorithm is a prediction model. It is quite usual such models are computationaly expensive and growth of the quality of such models leads to the deterioration in the inference speed. However it is not always tradeoff between quality vs speed. In this paper we show it is possible to speed up model by using additional memory without loosing prediction quality significantly for a novel boosted trees algorithm called CatBoost. The idea is to combine two approaches: training fewer trees and merging trees into kind of hash maps (DecisionTensors). The proposed method allows for pareto-optimal reduction of the computational complexity of the decision tree model with regard to the quality of the model. In the considered example number of lookups was decreased from 5000 to only 6 (speedup factor of 1000) while AUC score of the model was reduced by less than per mil.
\end{abstract}

@@

For example, Topological Trigger in LHCb is one of the places with very strict applying time restrictions. In [2] original MatrixNet [1] model allowed to significantly improve quality. But to get acceptable applying time the model was coarsed and converted to Bonsai Boosted Decision Tree format [4].

In this paper, we consider a modification of the training process of a predictive models by successor of MatrixNet, decision tree-based algorithm CatBoost [3]. But our approach is also applicable to other boosting decision tree-based algorithms.

@@

CatBoost is the implementation of a tree-based gradient boosting in which oblivious decision trees are used [3]. The values of each feature $i$ are discretized into baskets using the boundaries $B_i={b_{i1},...,b_{ip_i}}$, where $p_i$ is the number of boundaries. The boundaries themselves are determined in advance using various approaches.
Original feature vector with real values is replaced by a vector of binary values $0$ and $1$ ($false$ and $true$). The value of the $i$-th attribute $f_i$ is replaced by a binary vector $G_i={g_{i1},...,g_{ip_i}}$ where $g_{ij} = f_i > b_{ij}$. Finally, all $G_i$ are concatenated into one vector. Consecutively, the resuling decision will a make bunch of boolean checks instead of comparing some real feature value with the thresholds.

@@

\section*{References}
\numrefs{99}
\item Gulin A, Kuralenok I, Pavlov D 2011 {\it Winning The Transfer Learning Track of Yahoo!'s Learning To Rank Challenge with YetiRank.} (Yahoo! Learning to Rank Challenge) p 63--76
\item Likhomanenko T, Ilten P, Khairullin E, Rogozhnikov A, Ustyuzhanin A, Williams M 2015 {\it LHCb Topological Trigger Reoptimization} (Journal of Physics: Conference Series)
\item Dorogush A V, Gulin A, Gusev G, Kazeev N, Ostroumova Prokhorenkova L, Vorobev A {\it Fighting biases with dynamic boosting.} ([arXiv:1706.09516])
\item Gligorov V, Williams M 2013 {\it Efficient, reliable and fast high-level triggering using a bonsai boosted decision tree} (JINST 8 P02013 [arXiv:1210.6861])
\endnumrefs
